Towards a clearer causal question underlying the association between cancer and dementia

### 2021-11-03 04_reports/manuscript.Rmd
need to fix all references and (.) order; footnote on tables; add 0 decimal; bold eappendix
submitted manuscript!

**Background:** Several observational studies have described an inverse association between cancer diagnosis and dementia. Several biological mechanisms and sources of bias have been proposed. Since bias cannot be assessed without a clear causal question, we propose to study the controlled direct effect of the protein Pin1 on the risk of dementia. With this case-study, we will outline the needed assumptions and potential sources of bias and exemplify how these translate into the analytic decisions under the guidance of causal directed acyclic graphs. 
**Methods:** We used data from the Rotterdam Study, a population-based cohort. We illustrate the association between cancer diagnosis (as proxy for Pin1) and dementia diagnosis under two scenarios: (A) cancer proxy as "cancer ever vs. never"; (B) cancer diagnosis as time-varying proxy. We address confounding and censoring for death with inverse probability weights and contrast the complement of a weighted Kaplan-Meier survival estimator at 20-years of follow-up.
**Results:** Out of 3634 participants, 899 (25%) were diagnosed with cancer, of whom 53 (6%) had dementia, and 567 (63%) died. Among those without cancer, 15% (411) were diagnosed with dementia, and 667 (24%) died over follow-up. In Scenario A, the risk ratio (and 95%CI) after adjustment for confounding was 0.70 (0.49, 0.93); and with additional weights for death 0.91 (0.65, 1.19). In Scenario B, the risk ratio was 0.98 (0.71,1.19) and 1.05 (0.79, 1.29), respectively.
**Conclusion**: Being explicit about the underlying mechanism of interest and defining a clear causal contrast is key to maximize what we can learn about this association given available or readily-collected data, detecting and preventing potential biases.

@@

Many observational studies have consistently found that individuals with cancer have a lower risk of developing dementia when compared to individuals with no history of cancer [@ma2014;@hanson2016;@vanderwillik2018;@ospina2020]. These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015]. These research questions have led to discussions of repurposing or augmenting current cancer treatments, including chemotherapy, for dementia [@Snyder2017].

@@

In the observational setting, confounding could explain the observed association between Pin1 and dementia. In the causal diagram [@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and dementia diagnosis by time $t+k$ as $Y_{t+k}$. Both may share causes $L_1$, and to assess the causal relationship between $P_{t}$ and $Y_{t+k}$ we would thus adjust for $L_1$. Previous dementia studies have described age, sex, educational level and race/ethnicity as a minimal adjusting set of covariates[@ospina2020]. However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of dementia[@dementia_lancet] and should therefore be adjusted for. Although $L_1$ can be time-varying in nature, we only depict $L_1$ at one time-point for readability.

@@

Another challenge that arises when choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, i.e. the time when eligibility criteria are met, “treatment” is assigned, and screening for dementia would begin[@hernanInmortal2016].Emulating the eligibility criteria to participate in a trial for Pin1-targeting drugs will not necessarily be possible in a cohort setting recruiting participants for discordant reasons. Moreover, the latter may not align with the moment at which cancer diagnosis is measured since it happens continuously over follow-up and this situation may introduce immortal-time bias[@hernanInmortal2016]. For example, a study performed using data from the Framingham Heart Study [@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer" [@hanson2016]). This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis. By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of dementia diagnosis over follow-up can be defined as "ever cancer" [@anderson1983, @hernanInmortal2016].

This problem is partly mitigated by recognizing the time-varying nature of cancer diagnosis. Some studies have considered cancer diagnosis as a time-dependent exposure [@white2013; @hanson2016; @bowles2017]. The price we pay for this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible. The time between first biological changes that eventually can lead to cancer and cancer manifestation can range between five and forty years[@nadler2013]. Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up. Thus, in Figure 3 we included death prior to cancer diagnosis as $D_{t+1}$ and an arrow between $D_{t+1}$ and $P_{t+1}$ that represents a deterministic association such as that $P_{t+1}$ is only observed if $D_{t+1}$ is zero. In addition, we added an arrow between $L_1$ and $D_{t+1}$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease.

@@

To isolate the direct effect of $P_{t+1}$ in $Y_{t+k}$ through measurement of $C_{t+1}$ we have at least two alternatives of causal questions we can ask. One, we can imagine a causal question where we decompose the total effect of cancer by the different pathways that affect dementia and death separately[@stensrud2020]. Two, we can define a scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as an independent censoring event by relying on the assumption that we have measured all shared causes $L_2$ to block the pathway $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$. Previous studies have defined death as a censoring event [@frain2017], although failed to explicitly define  the independent censoring assumption and did not consider its plausibility. Moreover, adjusting for time-fixed shared causes between dementia and death may be insufficient to block this pathway, and therefore time-varying measurements of $L_2$ should be considered.

@@

On the other hand, death as a competing event is a challenge that has some straightforward strategies, beginning first and foremost by choosing the causal parameter (or estimand) of interest [@young2020, @rojas_medrxiv]. In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death due to cancer and other causes could have been prevented, yet without an explicit intervention, which makes it ambiguous. This is different than conceiving a drug-target that increases Pin1 expression only in brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer). As opposed to prior studies that implicitly address a direct effect, and that define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption [@weave2012, @vangeloven2014, @willems2018, @young2020] when the estimand of interest is the controlled direct effect. Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null. This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared (time-varying) causes of dementia and death are needed. In addition, presenting the proportion of participants that died prior to dementia diagnosis in each arm, as well as the proportion of participants in each status over time, improves transparency and puts in evidence the limitations of the data to answer this question.

Pin1 is only one of the several mechanisms proposed by other investigators as an explanation for the inverse cancer-dementia relationship. Certainly, cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression. To understand how – if at all – the cancer-dementia association informs the potential effects of any other mechanism or treatment strategy, and its connection to collected data, we may require different causal representations. For example, if researchers used this association to inform the possible effects of different chemotherapeutics on cognitive decline among patients undergoing treatment for cancer, different challenges would arise for mapping the observed association to the hypothetical randomized trial underlying that new research question. Notably, that target trial, unlike the one considered here, would include cancer diagnosis as part of the eligibility, and thus researchers would need to instead grapple with how using data on persons without cancer is useful and useable [@huitfeldt2016].
Each question requires thinking about bias anew and each question brings its own set of challenges and opportunities.

@@

### Table 1: Descriptive characteristics of individuals who had a cancer diagnosis and of those free of cancer diagnosis over follow-up.

@@

**Figure 1. Pin1 as biomarker and risk of dementia.** Causal directed acyclic graph where Pin1 at time $t$ is represented as $P_t$, Dementia at time $t+k$ is represented as $Y_{t+k}$. $L_1$ represents shared causes of Pin1 and Dementia, such as smoking. To isolate the effect of $P_t$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$. 
Gray nodes and arrows, which represent additional data-generating mechanisms, will be described progressively in the following figures.

@@

**Figure 2. Incident cancer diagnosis as proxy for Pin1 expression.** This causal graph is an expansion of Figure 1, includes Pin1 expression at time $P_{t+1}$. Incident cancer diagnosis is represented as $C_t$ and $C_{t+1}$. To isolate the effect of $P_t$ and $P_{t+1}$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$ and $Y_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$. Although we represent $L_1$ as a single node for readability, $L_1$ is time-varying too.

@@

**Figure 3. Time-varying cancer diagnosis and immortal-time bias.**  This causal graph is an expansion of Figure 2, where $D_{t+1}$ represents death at time $t+1$. Cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$

@@

**Figure 4. Death as competing event of dementia.** In this graph we only focus attention to the exposure as if it was measured for all at time $P_{t+1}$. We include $D_{t+k}$ as death at time $t+k$ since $Y_{t+1}$ is only observable when participants are alive at $t+k$. $L_2$ represents possible shared causes of dementia and death (such as cardiovascular comorbidities). Pin1 may affect the risk of death through cancer diagnosis, represented as an arrow between $C_{t+1}$ and $D_{t+k}$. To isolate the direct effect of $P_{t+1}$ and $Y_{t+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$ and $Y_{t+k} \leftarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$

@@

**Figure 5. Causal graph for effect of Pin1 through the proxy of cancer diagnosis in the risk of dementia, including all challenges discussed through Figures 1 to 4.**

>>>>

**Background:** Several observational studies have described an inverse association between cancer diagnosis and dementia. Several biological mechanisms and sources of bias have been proposed. Since bias cannot be assessed without a clear causal question, we propose to study the controlled direct effect of the protein Pin1 on the risk of dementia. With this case-study, we will outline the needed assumptions and potential sources of bias and exemplify how these translate into the analytic decisions under the guidance of causal directed acyclic graphs.

**Methods:**We used data from the Rotterdam Study, a population-based cohort. We estimate the association between cancer diagnosis (as proxy for Pin1) and dementia diagnosis using two different proxy methods and with confounding and censoring for death addressed with inverse probability weights. We estimate and compare the complements of a weighted Kaplan-Meier survival estimator at 20-years of follow-up.

**Results:** Out of 3634 participants, 899 (25%) were diagnosed with cancer, of whom 53 (6%) had dementia, and 567 (63%) died. Among those without cancer, 15% (411) were diagnosed with dementia, and 667 (24%) died over follow-up. Depending on the confounding and selection bias control, and the way in which cancer was used as a time-varying proxy, the risk ratio ranged from 0.70 (0.49, 0.93) to 1.05 (0.79, 1.29).

**Conclusion:** Being explicit about the underlying mechanism of interest and defining a clear causal contrast is key to maximize what we can learn about this association given available or readily-collected data, and to defining, detecting, and preventing potential biases.

@@

Many observational studies have consistently found that individuals with cancer have a lower risk of developing dementia when compared to individuals with no history of cancer[@ma2014;@hanson2016;@vanderwillik2018;@ospina2020]. These findings have motivated substantial research toward mechanistic explanations, including molecular and genetic pathways that may explain this association[@behrens2009;@harris2014;@nudelman2019;@Papin2020;@driverbiogeront2014;@olson2019;@li2021;@driverpin12015]. These research questions have led to discussions of repurposing or augmenting current cancer treatments, including chemotherapy, for dementia[@Snyder2017].

@@

In the observational setting, confounding could explain the observed association between Pin1 and dementia. In the causal diagram[@whatif2020] in Figure 1, we present Pin1 expression as $P_{t}$ and dementia diagnosis by time $t+k$ as $Y_{t+k}$. Both may share causes $L_1$, and to assess the causal relationship between $P_{t}$ and $Y_{t+k}$ we would thus adjust for $L_1$. Previous dementia studies have described age, sex, educational level and race/ethnicity as a minimal adjusting set of covariates[@ospina2020]. However, environmental and behavioral factors such as smoking may translate into Pin1 over-expression[@tan2010] and are also related to the development of dementia[@dementia_lancet] and should therefore be adjusted for. Although $L_1$ can be time-varying in nature, we only depict $L_1$ at one time-point for readability.

@@

Another challenge that arises when choosing cancer diagnosis as the proxy for Pin1 is defining the time zero, i.e. the time when eligibility criteria are met, “treatment” is assigned, and screening for dementia would begin[@hernanInmortal2016].Emulating the eligibility criteria to participate in a trial for Pin1-targeting drugs will not necessarily be possible in a cohort setting recruiting participants for discordant reasons. Moreover, the latter may not align with the moment at which cancer diagnosis is measured since it happens continuously over follow-up and this situation may introduce immortal-time bias[@hernanInmortal2016]. For example, a study performed using data from the Framingham Heart Study[@driver2012] defined the exposed group with cancer as those participants with prevalent or incident cancer diagnosis (alternatively defined as "ever cancer"[@hanson2016]). This meant that a participant who had cancer diagnosis over follow-up contributed all their person-time to the cancer arm, including the time prior to the cancer diagnosis. By defining the exposure after the time of inclusion to the cohort, only participants who remain alive and free of dementia diagnosis over follow-up can be defined as "ever cancer"[@anderson1983, @hernanInmortal2016].

This problem is partly mitigated by recognizing the time-varying nature of cancer diagnosis. Some studies have considered cancer diagnosis as a time-dependent exposure[@white2013; @hanson2016; @bowles2017]. The price we pay for this approximation is that implicitly, this means that Pin1 would over-express at the time of cancer diagnosis and not before, which is biologically implausible. The time between first biological changes that eventually can lead to cancer and cancer manifestation can range between five and forty years[@nadler2013]. Moreover, cancer diagnosis will only be measured in the subset of participants who are alive over follow-up. Thus, in Figure 3 we included death prior to cancer diagnosis as $D_{t+1}$ and an arrow between $D_{t+1}$ and $P_{t+1}$ that represents a deterministic association such as that $P_{t+1}$ is only observed if $D_{t+1}$ is zero. In addition, we added an arrow between $L_1$ and $D_{t+1}$, since covariates such as smoking may affect Pin1 over-expression but also affect the risk of death due to other causes such as from chronic obstructive pulmonary disease.

@@

To isolate the direct effect of $P_{t+1}$ in $Y_{t+k}$ through measurement of $C_{t+1}$ we have at least two alternatives of causal questions we can ask. One, we can imagine a causal question where we decompose the total effect of cancer by the different pathways that affect dementia and death separately[@stensrud2020]. Two, we can define a scenario where death could have been prevented. The latter is defined as the controlled direct effect, where death is considered as an independent censoring event by relying on the assumption that we have measured all shared causes $L_2$ to block the pathway $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$. Previous studies have defined death as a censoring event[@frain2017], although failed to explicitly define  the independent censoring assumption and did not consider its plausibility. Moreover, adjusting for time-fixed shared causes between dementia and death may be insufficient to block this pathway, and therefore time-varying measurements of $L_2$ should be considered.

@@

On the other hand, death as a competing event is a challenge that has some straightforward strategies, beginning first and foremost by choosing the causal parameter (or estimand) of interest[@young2020; @rojas_medrxiv]. In this study we chose the controlled direct effect, which represents the effect of Pin1 (or cancer) in a setting where death due to cancer and other causes could have been prevented, yet without an explicit intervention, which makes it ambiguous. This is different than conceiving a drug-target that increases Pin1 expression only in brain tissue, with no side effects that could increase cancer risk (and thus, death due to cancer). As opposed to prior studies that implicitly address a direct effect, and that define censoring as ignorable[@frain2017], we show how point estimates change substantially when we include weights for death to relax the independent censoring assumption[@weave2012; @vangeloven2014; @willems2018; @young2020] when the estimand of interest is the controlled direct effect. Bounds to assess extreme scenarios of dependence between death and dementia[@peterson1976] illustrate the wide range of possible point estimates that cross the null. This shows that even with the effort of adjusting for time-varying covariates, we may be far from meeting this assumption and thus better efforts to measure shared (time-varying) causes of dementia and death are needed. In addition, presenting the proportion of participants that died prior to dementia diagnosis in each arm, as well as the proportion of participants in each status over time, improves transparency and puts in evidence the limitations of the data to answer this question.

Pin1 is only one of the several mechanisms proposed by other investigators as an explanation for the inverse cancer-dementia relationship. Certainly, cancer diagnosis represents a complex and heterogeneous health condition that exceeds the representation of Pin1 expression. To understand how – if at all – the cancer-dementia association informs the potential effects of any other mechanism or treatment strategy, and its connection to collected data, we may require different causal representations. For example, if researchers used this association to inform the possible effects of different chemotherapeutics on cognitive decline among patients undergoing treatment for cancer, different challenges would arise for mapping the observed association to the hypothetical randomized trial underlying that new research question. Notably, that target trial, unlike the one considered here, would include cancer diagnosis as part of the eligibility, and thus researchers would need to instead grapple with how using data on persons without cancer is useful and useable[@huitfeldt2016].
Each question requires thinking about bias anew and each question brings its own set of challenges and opportunities.

@@

### Table 1: Descriptive characteristics of individuals included in this study.

<br>

@@

**Figure 1. Causal directed acyclic graph highlighting confounding of the potential effect of Pin1 on risk of dementia.** Pin1 at time $t$ is represented as $P_t$; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia, such as smoking. To isolate the effect of $P_t$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$. Gray nodes and arrows are described progressively in Figures 1-5.

@@

**Figure 2.  Causal directed acyclic graph highlighting the use of incident cancer diagnosis as a proxy for Pin1 expression.** Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$. To isolate the effect of $P_t$ and $P_{t+1}$ on $Y_{t+k}$ we need to block the backdoor path $Y_{t+k} \leftarrow L_1 \rightarrow P_t$ and $Y_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$. Although we represent $L_1$ as a single node for readability, $L_1$ is time-varying too. Gray nodes and arrows are described progressively in Figures 1-5.

@@

**Figure 3. Causal directed acyclic graph highlighting the time-varying nature of cancer diagnosis and immortal time bias.**  Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$. $D_{t+1}$ represents death at time $t+1$, cancer diagnosis at $t+1$ can only be measured among those who are alive at $t+1$. Gray nodes and arrows are described progressively in Figures 1-5.

@@

**Figure 4. Causal directed acyclic graph highlighting death as a competing event of dementia** Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$; $D_{t+1}$ represents death at time $t+1$. In this graph we only focus attention to the exposure as if it was measured for all at time $P_{t+1}$. We include $D_{t+k}$ as death at time $t+k$ since $Y_{t+1}$ is only observable when participants are alive at $t+k$. $L_2$ represents possible shared causes of dementia and death (such as cardiovascular comorbidities). Pin1 may affect the risk of death through cancer diagnosis, represented as an arrow between $C_{t+1}$ and $D_{t+k}$. To isolate the direct effect of $P_{t+1}$ and $Y_{t+k}$, we have to block the backdoor pathway from $Y_{t+k} \leftarrow L_2 \rightarrow D_{t+k} \leftarrow C_{t+1} \leftarrow P_{t+1}$ and $Y_{t+k} \leftarrow D_{t+k} \leftarrow L_1 \rightarrow P_{t+1}$. Gray nodes and arrows are described progressively in Figures 1-5.

@@

**Figure 5. Causal directed acyclic graph depicting multiple challenges to using the proxy of cancer diagnosis to study the effect of Pin1 on risk of dementia.** Pin1 at time $t$ and $t+1$ are represented as $P_t$ and $P_{t+1}$, respectively; dementia at time $t+k$ is represented as $Y_{t+k}$; $L_1$ represents shared causes of Pin1 and dementia; incident cancer diagnosis at time $t$ and $t+1$ are represented as $C_t$ and $C_{t+1}$; $D_{t+1}$ and $D_{t+k}$ represents death at time $t+1$ and $t+k$; $L_2$ represents possible shared causes of dementia and death. The distinct challenges were highlighted separately in Figures 1-4.
